WSClean: an implementation of a fast, generic wide-field imager

### 2014-01-30 articles/2014/Article-WSClean/wsclean.tex
Some progress

For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite). This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. TODO: discuss A-Projection (CASA manual say: ``A-Projection is an algorithm to account for the effects of the antenna primary beam (PB) during imaging. The time-dependent effects of the PB are projected-out during the imaging phase and the PB is included in the prediction phase of the iterative image deconvolution (see Bhatnagar, Cornwell, Golap \& Uson 2008, A\&A, 487, 419'') The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is an efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms combined with snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. This work improves the field of wide-field imaging in the following ways:
\begin{itemize}
 \item The performance of a w-stacking implementation is compared to a w-projection implementation.
 \item The advantages of using the w-snapshots algorithm are analysed.
 \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
 \item A new implementation is released, which we believe to be the fastest generic widefield imager available for many imaging scenarios.
\end{itemize}

The w-stacking algorithm is described in Sect.~\ref{sec:wstacking}. Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}. Using various MWA observations, including observations with varying elevation, the performance will be analysed in Sect.~\ref{sec:analysis}.

@@

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion, replace with deeper images.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith. As shown, all-sky imaging can be quite efficient with this technique. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with major iterations similar to a non-shifted set.

@@

\subsection{Accuracy}
\begin{table}%
\caption{Results on imaging accuracy measurements.} \label{tbl:accuracy-measurements}%
\begin{center}\begin{tabular}{llll}%
\hline\hline
& \textbf{WSClean} & \textbf{WSClean}    & \textbf{CASA} \\
&                  & \textbf{+ recentre} & \\
\hline
Source flux standard error & 1.31\% & & 1.34\% \\
RMS in residual image & 0.94 mJy & & 1.90 mJy \\
Computational time & 8 min & & forever (TODO) \\
TODO at ZA=10\degree & & \\
\hline\hline
\end{tabular}\end{center}\end{table}

>>>>

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite). This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is the most efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. This work improves the field of wide-field imaging in the following ways:
\begin{itemize}
 \item The performance of w-stacking is compared to w-projection.
 \item Parameters that affect the w-snapshots algorithm are analysed.
 \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
 \item A new implementation is released, which we believe to be the fastest generic widefield imager available.
\end{itemize}

The w-stacking algorithm is described in Sect.~\ref{sec:wstacking}. Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}. The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.

@@

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion, replace with deeper images.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith. As shown, all-sky imaging can be quite efficient with this technique. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.

@@

\subsection{Accuracy}
\begin{table}%
\caption{Results on imaging accuracy measurements.} \label{tbl:accuracy-measurements}%
\begin{center}\begin{tabular}{llll}%
\hline\hline
& \textbf{WSClean} & \textbf{WSClean}    & \textbf{CASA} \\
&                  & \textbf{+ recentre} & \\
\hline
\multicolumn{1}{l}{\textit{Zenith angle 0\degree}} \\
\hline
Source flux standard error & 1.31\% & & 1.34\% \\
RMS in residual image & 0.94 mJy & & 1.90 mJy \\
Computational time & 8 min & & 19.3 min \\
\hline
\multicolumn{1}{l}{\textit{Zenith angle 10\degree}} \\
\hline
Source flux standard error & 1.75\% & & 2.41 \%\\
RMS in residual image & 0.88 mJy & & 1.04 mJy \\
Computational time & 15.3 min & & 178.2 min \\
\hline\hline
\end{tabular}\end{center}\end{table}
